# UCI Salmon Fishery Data Sources {.unnumbered}

---
execute:
  echo: false
date: "`r Sys.Date()`"
format:
  html:
    code-fold: true
    code-tools: true
    code-summary: "Show the code"
---

```{r, echo = F, message = F}

#| warning: false
#| message: false


# clear environment
rm(list=ls())

# load packages
library(tidyverse)
library(janitor)
library(scales)
library(magrittr)
```

# Data Sources

```{r echo = F}


# harvest trends datasets 1966-2015
# https://catalog.epscor.alaska.edu/dataset/commercial-recreational-and-personal-use-harvest-trends-in-the-kenai-river-salmon-fishery-1966-2015

# overall run trends datasets 1976-2015
# https://catalog.epscor.alaska.edu/dataset/total-run-sizes-of-chinook-and-sockeye-salmon-in-the-kenai-river-1976-2015

```

Note on data sources for Northern Kenai Peninsula Fisheries Harvest, from Schoen et al. 2017 supplemental materials:

"Harvest and effort data are reported differently for the commercial, recreational, and personal-use fisheries in Upper Cook Inlet, such that it is not possible to calculate the harvest and effort of salmon produced specifically by the Kenai River (including its tributaries) by each fishing sector. To make comparisons within a common geographic area, we aggregated harvest and effort data from the northern Kenai Peninsula, ranging from the Kasilof River in the south to Ingram Creek in the north. The Kenai River is the predominant salmon-producing river in this area.

We aggregated harvest and effort data from the following fisheries:

-   Commercial: Central District drift gill net and east-side set gill net (Shields and Dupuis 2016);

-   Recreational: Northern Kenai Peninsula Management Area (Begich et al. 2013);

-   Personal-use: Kenai River dip net, Kasilof River dip net, and Kasilof River set net (Fall et al. 2015; Shields and Dupuis 2016). ADF&G also reports harvest and effort for an “unknown” personal-use fishery in Upper Cook Inlet. This harvest and effort was reported on permits on which the fishery was left blank. We allocated this harvest and effort to each known fishery based on the proportions of accurately reported harvest and effort in each year."

<br>

## Read in data

```{r echo = F}

### commercial

# read in comm fish data
dir <- "other/input/harvest_trends/uci-commercial-salmon-harvest.csv"
uci_comm <- read.csv(dir) %>%
  clean_names() %>%
  # remove percentage columns
  select(-contains(("_2"))) %>%
  pivot_longer(cols = c("drift",
                        "essn",
                        "wssn",
                        "nsn"), names_to = "fishery", values_to = "count") %>%
  # consolidate counts from various fisheries
  group_by(species,year) %>%
  summarise(total_count = sum(count)) %>%
  mutate(harvest_group = "Commercial") %>%
  transform(year = as.character(year))

# the commercial harvest time series is the longest, so we will make that the axis extent for the other figures too
comm_year_min <- as.numeric(min(uci_comm$year))
comm_year_max <- as.numeric(max(uci_comm$year))

<<<<<<< HEAD
# for some visualization purposes, we want the y-axis to show the scale of the largest harvest category, commercial. Species are aggregated
# comm_harvest_min <- group_by(year) %>% summarise(total_comm_harvest = sum(total_ct)) # throws error
#comm_harvest_max
=======
#################
>>>>>>> 035abfad03e65f915a08ba296ae0255bbd2f12a3

### personal use

<<<<<<< HEAD
<br>

## Personal Use Harvest

```{r echo = F}
=======
>>>>>>> 035abfad03e65f915a08ba296ae0255bbd2f12a3
# read in pre 1996 PU
dir <- "other/input/harvest_trends/puharvest_pre1996.csv"
puharvest_pre1996 <- read.csv(dir) %>%
  clean_names() %>%
  # prep format
  pivot_longer(cols = c("harvest_sockeye",
                        "harvest_chinook",
                        "harvest_coho",
                        "harvest_chum",
                        "harvest_pink"), names_to = "species", values_to = "count") %>%
  select(-harvest_total) %>%
  # corrected species names
  mutate(species = case_when(
    species == "harvest_sockeye" ~ "Sockeye",
    species == "harvest_chinook" ~ "Chinook",
    species == "harvest_chum" ~ "Chum",
    species == "harvest_pink" ~ "Pink",
    species == "harvest_coho" ~ "Coho")) %>%
  # consolidate data among fisheries, as the figure does not segregate them
  group_by(species, year) %>%
  summarise(total_count = sum(count)) %>%
  transform(year = as.character(year))


# read in post 1996 PU
dir <- "other/input/harvest_trends/puharvest_post1996.csv"
puharvest_post1996 <- read.csv(dir) %>%
  clean_names() %>%
  select(-source,-page) %>%
  # remove commas
  mutate_all(~sub(",","",.)) %>%
  # prep format
  pivot_longer(cols = c("fish_cr",
                        "kasilof_set",
                        "kasilof_dip",
                        "kenai",
                        "unknown"), names_to = "fishery", values_to = "count") %>%
  transform(count = as.numeric(count)) %>%
  select(-fishery) %>%
  # replace all 0 with NA
  mutate(count = na_if(count,0)) %>%
  group_by(species, year) %>%
  # recall to include na.rm = T
  summarise(total_count = sum(count, na.rm = T)) 

# combine pre-1996 and post-1996 PU data
uci_pu <- bind_rows(puharvest_pre1996,puharvest_post1996) %>%
  mutate(harvest_group = "Personal Use") 

#########

### sport

<<<<<<< HEAD
# add secondary axis


# suggested alternative: secondary chart below w/ household days





# number format on y axis
# capitalize legend
# reverse stack order
# match colors
# secondary y-axis

```

<br>

## Sport Harvest

```{r echo = F}
=======
>>>>>>> 035abfad03e65f915a08ba296ae0255bbd2f12a3
# read in uci sport fish data
dir <- "other/input/harvest_trends/sportharvest.csv"
uci_sport <- read.csv(dir) %>%
  clean_names() %>%
  select(contains(c("year","area","salmon","trout","dolly","grayling","pike","other"))) %>%
  # consolidate all non-salmon species into "other"
  rowwise() %>%
  mutate(other = sum(rainbow_trout, 
                     lake_trout, 
                     dolly_varden,
                     arctic_grayling,
                     northern_pike,
                     other,
                     na.rm = T)) %>%
  # remove non-salmon species columns
  select(-rainbow_trout, -lake_trout, -dolly_varden,-arctic_grayling,-northern_pike) %>%
  pivot_longer(cols = contains(c("salmon","other")),
               names_to = "species",
               values_to = "count") %>%
  # consolidate counts from various fisheries into overall sport fishery
  group_by(year,species) %>%
  summarise(total_count = sum(count)) %>%
  # fix species names
  mutate(species = case_when(
    species == "chinook_salmon" ~ "Chinook",
    species == "chum_salmon" ~ "Chum",
    species == "coho_salmon" ~ "Coho",
    species == "pink_salmon" ~ "Pink",
    species == "sockeye_salmon" ~ "Sockeye",
    species == "other" ~ "Other")) %>%
  # name user group
  mutate(harvest_group = "Sport") %>%
  transform(year = as.character(year))

#### combine commercial, PU, and sport into single dataframe

# combine dataframes for comm, pu, and sport
dat <- bind_rows(uci_comm, uci_pu, uci_sport)


```


# plots

```{r echo = F, message = F}

# horizontal faceted plot
dat %>%
  transform(year = as.numeric(year)) %>%
  filter(year < 2023) %>%
  group_by(year,harvest_group) %>%
  summarise(total_harvest = sum(total_count)) %>%
  ggplot() +
  geom_bar(aes(x = year, y = total_harvest, fill = harvest_group), stat = "identity") +
  scale_y_continuous(labels = label_comma()) +
  theme_classic() +
  xlab("") +
  ylab("Total Harvest") +
  ggtitle("Upper Cook Inlet Salmon Harvest (All Species)") 



# report/include numbers for educational / traditional use; make point not big enough to plot here

# plots of value / size



```




## Commercial Harvest

```{r echo = F}
# comm plot
comm_fig <- uci_comm %>%
  ggplot() +
  geom_bar(aes(x = as.numeric(year), y = total_count, fill = species), stat = "identity", position = position_stack(reverse = T)) +
  scale_y_continuous(labels = label_comma()) +
  xlab("") +
  ylab("Commercial Harvest") +
  theme_classic() +
  ggtitle("Upper Cook Inlet Commercial Harvest")
comm_fig


# need clarification on and source for effort: number of permits fished. where; both setnet and drift?? check against shields and dupuis and find parallel in lipke & stumpf 2024
```

<br>


## Personal Use Harvest



```{r echo = F}

# PU plot
pu_fig <- uci_pu %>%
  filter(species != "HouseholdDaysFished") %>%
  ggplot() +
  geom_bar(aes(x = as.numeric(year), y = total_count, fill = species),  stat = "identity", position = position_stack(reverse = T)) +
  scale_y_continuous(labels = label_comma()) +
  xlab("") +
  ylab("Personal Use Harvest") +
  xlim(comm_year_min,comm_year_max) +
  theme_classic()
 #pu_fig

# add secondary axis


# 1) get effort data in new column
uci_pu2 <- uci_pu %>% filter(species == "HouseholdDaysFished") %>%
  select(-species)
colnames(uci_pu2) <- c("year","household_days_fished","harvest_group")
uci_pu2 <- left_join(uci_pu,uci_pu2) %>%
  filter(species != "HouseholdDaysFished")


# 2) plot PU fig v2
# PU plot
scale_factor <- 16

pu_fig2 <- uci_pu2 %>%
  filter(species != "HouseholdDaysFished") %>%
  ggplot() +
  geom_bar(aes(x = as.numeric(year), y = total_count, fill = species),  stat = "identity", position = position_stack(reverse = T)) +
  geom_line(aes(x = as.numeric(year), y = household_days_fished * scale_factor), color = "black", size = 1.05) +
  scale_y_continuous(name = "Personal Use Harvest", 
                     sec.axis = sec_axis(~ ., labels = number_format(scale=1/scale_factor), name="Effort (Household Days Fished)"),
                     labels = comma) +
  xlab("") +
  ylab("Personal Use Harvest") +
  xlim(comm_year_min,comm_year_max) +
  theme_classic() +
  ggtitle("Northern Kenai Peninsula Personal Use Salmon Fishery")
pu_fig2

# capitalize legend
# reverse stack order
# match colors

# see how this looks as a faceted plot with effort on top ?

# how does this plot look against sc ak population trend?
```






<br>

## Sport Harvest


```{r echo = F}
# sport plot
sport_fig <- uci_sport %>%
  ggplot() +
  geom_bar(aes(x = as.numeric(year), y = total_count, fill = species), stat = "identity",position = position_stack(reverse = T)) +
  scale_y_continuous(labels = comma) +
  xlab("") +
  ylab("Sport Harvest") +
  xlim(comm_year_min,comm_year_max) +
  ylim(0,800000) +
  theme_classic()
#sport_fig


# plot sport v2 with effort

# add effort data
sport_effort <- read.csv("other/input/harvest_trends/sportharvest_effort.csv") %>%
  clean_names() %>%
  select(year,kenai_river_total) %>%
  rename(angler_days = kenai_river_total) %>%
  # remove commas
  mutate_all(~sub(",","",.)) %>%
  transform(angler_days = as.numeric(angler_days),
            year = as.character(year))
# join
uci_sport2 <- left_join(uci_sport,sport_effort)

# plot
scale_factor <- 2

sport_fig2 <- uci_sport2 %>%
  ggplot() +
  geom_bar(aes(x = as.numeric(year), y = total_count, fill = species),  stat = "identity", position = position_stack(reverse = T)) +
  geom_line(aes(x = as.numeric(year), y = angler_days * scale_factor), color = "black", size = 1.05) +
  scale_y_continuous(name = "Sport Harvest", 
                     sec.axis = sec_axis(~ ., labels = number_format(scale=1/scale_factor), name="Effort (Angler Days)"),
                     labels = comma) +
  xlab("") +
  ylab("Sport Harvest") +
  xlim(comm_year_min,comm_year_max) +
  scale_y_continuous(labels = label_comma()) +
  theme_classic() +
  ggtitle("Upper Cook Inlet Freshwater Sport Fisheries")
sport_fig2

# y axis for numberf of angler days is missing
```

Next:

-   find commercial effort data
-   map fig(s) to indicate where data is sourced from
-   match stack colors to original fig
-   match stack order to legend order
-   explanatory text (comm has longest time series, etc)
-   diff colors for fig 1 than any other used
-   representative pictures of each fishery

Idea for NEW overall fig - stacked bar plot of harvest by fishery (& sub stack by species?) superimposed with total harvest ... can visualize remaining escapement in context of harvest. --\> might be hard to do becuase the above figures do not include solely kenai fish

<<<<<<< HEAD
-   secondary axes for effort: maybe? https://johnmackintosh.net/blog/2022-03-13-dual-axis/
=======
>>>>>>> 035abfad03e65f915a08ba296ae0255bbd2f12a3

-   step 1: recreate fig as is

<<<<<<< HEAD
-   step 2: show proportional sizes of figs in context of comm
=======
- place citations/sources for data below each plot


See p 91 in lipka & stumpf 2024, all uci sockeye harvests combined

plan - draft for sport and pu; send to anna, as for effort data for comm


>>>>>>> 035abfad03e65f915a08ba296ae0255bbd2f12a3

-   step 3: show stacked proportion by fishery by year
